An Extendable \proglang{Python} Implementation of Robust Optimisation Monte Carlo

### 2022-12-21 article/latex_files/ch4/4-2-inference.tex
fixed spaces and commas

\begin{Code}
---------------------------------- python ----------------------------------
>>> romc.sample(n2)
----------------------------------------------------------------------------
\end{Code}

@@

\begin{Code}
---------------------------------- python ----------------------------------
>>> romc.compute_expectation(h) # E[h(x)|theta]
>>> romc.eval_unorm_posterior(theta, eps_cutoff=False) # unnorm p(theta)
>>> romc.eval_posterior(theta, eps_cutoff=False) # normalised p(theta)
----------------------------------------------------------------------------
\end{Code}

@@

\begin{Code}
------------------------------ python snippet ------------------------------  
  # Inference part
  seed = 21
  n2 = 50
  romc.sample(n2=n2, seed=seed)

  # visualize region, adding the samples now
  romc.visualize_region(i=1)

  # Visualise marginal (built-in ELFI tool)
  romc.result.plot_marginals(weights=romc.result.weights,
                             bins=100, range=(-3,3))

  # Summarize the samples (built-in ELFI tool)
  romc.result.summary()
  # Number of samples: 19300
  # Sample means: theta: -0.0116

  # compute expectation
  exp_val = romc.compute_expectation(h=lambda x: np.squeeze(x))
  print("Expected value   : %.3f" % exp_val)
  # Expected value   : -0.012

  exp_var = romc.compute_expectation(h=lambda x: np.squeeze(x)**2)
  print("Expected variance: %.3f" % exp_var)
  # Expected variance: 1.120

  # eval unnorm posterior
  romc.eval_unnorm_posterior(theta=0)

  # check eval posterior
  romc.eval_posterior(theta=0)
----------------------------------------------------------------------------
\end{Code}

% \begin{figure}[h]
%     \begin{center}
%       \includegraphics[width=0.48\textwidth]{./latex_files/images/chapter3/example_marginal.png}
%     \end{center}
%   \caption[Histogram of the obtained samples at the 1D example.]{(a) Left: Histogram of the obtained samples. (b) Right: Acceptance region around \(\theta_1^*\) with the obtained samples plotted inside.}
%   \label{fig:example_sampling}
% \end{figure}

% \begin{figure}[h]
%     \begin{center}
%       \includegraphics[width=0.6\textwidth]{./latex_files/images/chapter3/example_posterior.png}
%     \end{center}
%   \caption[Approximate posterior evaluation, at the 1D example.]{Approximate posterior evaluation.}
%   \label{fig:approx_posterior}
% \end{figure}

>>>>

\begin{Code}
romc.sample(n2)
\end{Code}

@@

\begin{Code}
romc.compute_expectation(h) # E[h(x)|theta]
romc.eval_unorm_posterior(theta, eps_cutoff = False) # unnorm p(theta)
romc.eval_posterior(theta, eps_cutoff = False) # normalised p(theta)
\end{Code}

@@

\begin{Code}
# Inference part
seed = 21
n2 = 50
romc.sample(n2 = n2, seed = seed)

# visualize region, adding the samples now
romc.visualize_region(i = 1)

# Visualise marginal (built-in ELFI tool)
weights = romc.result.weights
romc.result.plot_marginals(weights = weights, bins = 100, range = (-3,3))

# Summarize the samples (built-in ELFI tool)
romc.result.summary()
# Number of samples  : 19300
# Sample means: theta: -0.0116

# compute expectation
exp_val = romc.compute_expectation(h = lambda x: np.squeeze(x))
print("Expected value   : %.3f" % exp_val)
# Expected value: -0.012

exp_var = romc.compute_expectation(h = lambda x: np.squeeze(x)**2)
print("Expected variance: %.3f" % exp_var)
# Expected variance: 1.120

# eval unnorm posterior
romc.eval_unnorm_posterior(theta = 0)

# check eval posterior
romc.eval_posterior(theta = 0)
\end{Code}

### 2022-12-21 article/latex_files/ch4/4-3-evaluation.tex
fixed references
fixed spaces and commas

\begin{Code}
---------------------------------- python ----------------------------------  
>>> romc.compute_divergence(gt_posterior, bounds, step, distance)
>>> romc.compute_ess()
----------------------------------------------------------------------------
\end{Code}

@@

The method \code{compute_ess} computes the Effective Sample Size (ESS)
using the following expression~\cite{Sudman1967},

@@

\begin{Code}
------------------------------ python snippet ------------------------------  
  # Evaluation part
  res = romc.compute_divergence(wrapper, distance="Jensen-Shannon")                                 
  print("Jensen-Shannon divergence: %.3f" % res)
  # Jensen-Shannon divergence: 0.035

  nof_samples = len(romc.result.weights)
  ess = romc.compute_ess()
  print("Nof Samples: %d, ESS: %.3f" % (nof_samples, ess))
  # Nof Samples: 19300, ESS: 16196.214
----------------------------------------------------------------------------  
\end{Code}

>>>>

\begin{Code}
romc.compute_divergence(gt_posterior, bounds, step, distance)
romc.compute_ess()
\end{Code}

@@

The method \code{compute_ess} computes the Effective Sample Size
(ESS) using the following expression,

@@

\begin{Code}
# Evaluation part
res = romc.compute_divergence(wrapper, distance = "Jensen-Shannon")
print("Jensen-Shannon divergence: %.3f" % res)
# Jensen-Shannon divergence: 0.035

nof_samples = len(romc.result.weights)
ess = romc.compute_ess()
print("Nof Samples: %d, ESS: %.3f" % (nof_samples, ess))
# Nof Samples: 19300, ESS: 16196.214
\end{Code}

### 2022-12-21 article/latex_files/ch4/4-4-extensibility.tex
fixed spaces and commas

\begin{Code}
---------------------------------- python ----------------------------------
class CustomOptim(OptimisationProblem):
    def __init__(self, **kwargs):
        super(CustomOptim, self).__init__(**kwargs)

@@

        # create local surrogate model as a function of theta
        def create_local_surrogate(model):
            return partial(local_surrogate, model_scikit=model)

@@

            # train Neural Network
            mlp = MLPRegressor(hidden_layer_sizes=(10,10), solver='adam')
            model = Pipeline([('linear', mlp)])
            model = model.fit(x, y)
            local_surrogates.append(create_local_surrogate(model))

        self.local_surrogates = local_surrogates
        self.state["local_surrogates"] = True
----------------------------------------------------------------------------
\end{Code}

@@

\begin{Code}
---------------------------------- python ----------------------------------
def solve_gradients(self, **kwargs):
  # custom solution procedure
  x = ...
  y = ...
  jac = ...
  hess_inv = ...

@@

\begin{Code}
---------------------------------- python ----------------------------------
class RomcOpimisationResult:
    def __init__(self, x_min, f_min, hess_appr):
        Parameters
        ----------
        x_min: np.ndarray (D,) or float, the minimum point
        f_min: float, distance at the minimum point
        hess_appr: np.ndarray (DxD), Hessian approximation at x_min
        """
----------------------------------------------------------------------------    
class NDimBoundingBox:
    def __init__(self, rotation, center, limits, eps_region):
        Parameters
        ----------
        rotation: np.array (D,D),  rotation matrix for the bounding box
        center: np.array (D,) center of the bounding box
        limits: np.ndarray, shape: (D,2), the limits of the bounding box
        eps_region: float, distance threshold 
----------------------------------------------------------------------------    
\end{Code}

>>>>

\begin{Code}
class CustomOptim(OptimisationProblem):
    def __init__(self, **kwargs):
        super(CustomOptim, self).__init__(**kwargs)

@@

        # create local surrogate model as a function of theta
        def create_local_surrogate(model):
            return partial(local_surrogate, model_scikit = model)

@@

            # train Neural Network
            mlp = MLPRegressor(hidden_layer_sizes = (10,10), solver = 'adam')
            model = Pipeline([('linear', mlp)])
            model = model.fit(x, y)
            local_surrogates.append(create_local_surrogate(model))

        self.local_surrogates = local_surrogates
        self.state["local_surrogates"] = True
\end{Code}

@@

\begin{Code}
----------------------------------------------------------------------------
def solve_gradients(self, **kwargs):
    # custom solution procedure
    x = ...
    y = ...
    jac = ...
    hess_inv = ...

@@

\begin{Code}
----------------------------------------------------------------------------
class RomcOpimisationResult:
    def __init__(self, x_min, f_min, hess_appr):
        Parameters
        ----------
        x_min: np.ndarray (D,) or float, the minimum point
        f_min: float, distance at the minimum point
        hess_appr: np.ndarray (DxD), Hessian approximation at x_min
        """
----------------------------------------------------------------------------
class NDimBoundingBox:
    def __init__(self, rotation, center, limits, eps_region):
        Parameters
        ----------
        rotation: np.array (D,D),  rotation matrix for the bounding box
        center: np.array (D,) center of the bounding box
        limits: np.ndarray, shape: (D,2), the limits of the bounding box
        eps_region: float, distance threshold
----------------------------------------------------------------------------
\end{Code}

### 2022-12-21 article/latex_files/ch5/5-ma2.tex
fixed spaces and commas

\noindent
The r.v. \(w_{k} \sim \mathcal{N}(0,1) \) is white noise and the two
parameters of interest, \(\theta_1, \theta_2\), model the dependence
from the previous observations. The number of sequential observations
\(T\) is a constant and set to \(T=100\). For securing
that the inference problem is identifiable, i.e.\ the likelihood has
only one mode, we use the prior proposed by~\cite{Marin2012},

@@

In order to show all the capabilities of the implementation, we
perform the inference (i) using the gradient-based optimizer, (ii)
using the Bayesian Optimisation scheme and (iii) fitting a Neural
Network (NN) as a surrogate model. We use the Rejection ABC algorithm
for evaluating all approaches. Replacing the typical quadratic
surrogate model with a NN serves as an illustrator of the
extensibility of our implementation. The replacement is done by coding
a custom optimisation function with a surrogate model of our own
preference, as shown in Chapter~\ref{subsec:extensibility}. For the
definition of the NN, we use the \pkg{MLPRegressor} class of the
\pkg{scikit-learn} package. Therefore, the NN substitutes the real
distance function \(d_i\) inside the proposal region
\(q_i~ \forall i\) at the inference phase i.e.~sampling,
computing an expectation and evaluating the posterior. In our example
we use a neural network of two hidden layers of 10 neurons each and we
train it sampling 500 examples from each proposal region.

@@

As stated above, ROMC is a ridiculously parallelisable
method. Therefore, it is straightforward to parallelise the fitting
part, i.e. (i) solving the optimisation problems and (ii) estimating
the proposal regions. Our implementation supports exploiting all the
available CPU cores through the built-in \proglang{Python} package
\pkg{multiprocess}\footnote{https://docs.python.org/3/library/multiprocessing.html}. In
Figure~\ref{fig:exec_parallel} we observe the execution times for
performing the inference on the MA2 model; the parallel version
performs both tasks almost five times faster than the sequential. The
result is reasonable given that the experiments have run in a single
machine with the Intel® Core™ i7-8750H Processor, which has six
separate cores.

>>>>

\noindent
The r.v. \(w_{k} \sim \mathcal{N}(0,1) \) is white noise and the two
parameters of interest, \(\theta_1, \theta_2\), model the dependence
from the previous observations. The number of sequential observations
\(T\) is a constant and set to \(T=100\). For securing
that the inference problem is identifiable, i.e., the likelihood has
only one mode, we use the prior proposed by~\cite{Marin2012},

@@

In order to show all the capabilities of the implementation, we
perform the inference (i) using the gradient-based optimizer, (ii)
using the Bayesian Optimisation scheme and (iii) fitting a Neural
Network (NN) as a surrogate model. We use the Rejection ABC algorithm
for evaluating all approaches. Replacing the typical quadratic
surrogate model with a NN serves as an illustrator of the
extensibility of our implementation. The replacement is done by coding
a custom optimisation function with a surrogate model of our own
preference, as shown in Chapter~\ref{subsec:extensibility}. For the
definition of the NN, we use the \pkg{MLPRegressor} class of the
\pkg{scikit-learn} package. Therefore, the NN substitutes the real
distance function \(d_i\) inside the proposal region
\(q_i~ \forall i\) at the inference phase, i.e., sampling,
computing an expectation and evaluating the posterior. In our example
we use a neural network of two hidden layers of 10 neurons each and we
train it sampling 500 examples from each proposal region.

@@

As stated above, ROMC is a ridiculously parallelisable
method. Therefore, it is straightforward to parallelise the fitting
part, i.e., (i) solving the optimisation problems and (ii) estimating
the proposal regions. Our implementation supports exploiting all the
available CPU cores through the built-in \proglang{Python} package
\pkg{multiprocess}\footnote{https://docs.python.org/3/library/multiprocessing.html}. In
Figure~\ref{fig:exec_parallel} we observe the execution times for
performing the inference on the MA2 model; the parallel version
performs both tasks almost five times faster than the sequential. The
result is reasonable given that the experiments have run in a single
machine with the Intel® Core™ i7-8750H Processor, which has six
separate cores.
